---
title: "Delta statistic"
format: html
---

```{r}
# HAV Clade IIIA variable clustering test
# Delta statistic/permutation test for IIIA sequences passing QC
# Also includes sanity check, missing data investigation
# Author: Mary Godec
# Date: July 2025

niter=50000
nburn=500
nperm=100
alpha=0.1
beta=0.05

#########################################################
# install required packages with Bioconductor
pkglist = c("dplyr", "ape", "phangorn", "phytools","finalfit", "ggtree")
lapply(pkglist, require, character.only = TRUE)

OUTPUT_DIR=Sys.getenv("OUTPUT_DIR")
METADATA_FILE=Sys.getenv("METADATA_FILE")
RENAMED_DELTA_WGS_TREE=Sys.getenv("RENAMED_DELTA_WGS_TREE")
DELTA_CODE=Sys.getenv("DELTA_CODE")

# https://github.com/mrborges23/delta_statistic
source(DELTA_CODE)
```

```{r}

meta <- read.csv(METADATA_FILE, header=TRUE)
unique(meta$homelessness)
unique(meta$illicit_ever)
tree <- ape::read.tree(RENAMED_DELTA_WGS_TREE)

## for other trees - drop outgroup, contextual IIIA, and HepA/USA/MA/0046/2018 (if in tree)
# tree <- drop.tip(tree, c("AB279732.1", "MN062167.1","ON524418.1","HepA/USA/MA/0046/2018"))

# # # option to recode homelessness and illicit_ever, similar to downstream JUNIPER analysis
# ## Encode ALL values so we don't have to drop tips - also lump empty/nan values into "No/Unknown"
md <- meta %>%
  mutate(
    homelessness = ifelse(is.na(homelessness) | homelessness %in% c("No", "Unknown", ""), "No/Unknown", homelessness),
    illicit_ever = ifelse(is.na(illicit_ever) | illicit_ever %in% c("No", "Unknown", ""), "No/Unknown", illicit_ever))

# filter metadata to only tips in tree 
tipnames <- tree$tip.label
# reorder to match order of tipnames
subs <- md[match(tipnames, md$MASPHL_Sequencing_ID ),]
nrow(subs)
```

```{r}
# subset/clean trees
print(is.rooted(tree))
# force polytomies to bifurcations
uditree <- multi2di(tree)
print(is.binary(uditree))

# We now have 0-length branches resulting from forcing bifurcations
# but for Delta, need tree to be metric (non-zero branch/edge lengths) 
print(summary(uditree$edge.length))
print((0.000000001 / mean(uditree$edge.length)) * 100) # What percentage is 0.000000001 of the mean?
# v small, likely won't affect branch length dependent calculations 

# convert 0 edge lengths to a tiny positive value that satisfies the branch_lengths > 0 requirement without influencing your δ results in any meaningful way
uditree$edge.length <- uditree$edge.length + 0.000000001

print(is.rooted(uditree))

# make sure all node labels (bootstrap values) are empty strings instead of missing - can also assign arbitrary bootstrap value if desired since this implementation doesn't consider bootstrap
# Especially important for ape - IQtree leaves the first label (for the root node) empty (seen as the leading blank in , NA, NA, NA...) - can cause off-by-one issues in how ape reads and stores them
uditree$node.label[is.na(uditree$node.label) | uditree$node.label == "" | uditree$node.label == NA ] <- ""

```

```{r}
# # drop all tips in tree that don't have yes or no in trait
hsubs <- subset(subs, subs$homelessness %in% c("Yes","No/Unknown"))
isubs <- subset(subs, subs$illicit_ever %in% c("Yes","No/Unknown"))


htrait <- hsubs$homelessness
itrait <- isubs$illicit_ever

h_factor <- as.numeric(as.character(factor(htrait, levels = c("No/Unknown", "Yes"), labels = c(1, 2))))
i_factor <- as.numeric(as.character(factor(itrait, levels = c("No/Unknown", "Yes"), labels = c(1, 2))))


# subset trees - drop all tips not in hsubs or isubs
htree <- drop.tip(uditree, tipnames[!tipnames %in% hsubs$MASPHL_Sequencing_ID])
itree <- drop.tip(uditree, tipnames[!tipnames %in% isubs$MASPHL_Sequencing_ID])

print(table(htrait))
print(table(itrait))
```


```{r}
#################
# l0: rate parameter of the exponential prior distribution
# se: standard deviation of the proposal distribution 
# sim: number of iterations
# thin: keeping only 10th iteration
# burn: 100 iterates are burned-in

#a:  alpha parameter (beta likelihood)
#b:  beta parameter (beta likelihood)
#x:  node entropies
#SYMdeltaA <- delta(calcfactor,calctree,0.1,0.0589,10000,10,100)

delta_plot_p <- function(calcfactor, calctree, titlestring) {
  SYMdeltaA <- delta(calcfactor, calctree, alpha, beta, niter, 10, nburn)
  print("deltaA")
  print(SYMdeltaA)
  random_delta <- rep(NA, nperm)
  for (i in 1:nperm) {
    rtrait <- sample(calcfactor)
    random_delta[i] <- delta(rtrait, calctree, alpha, beta, niter, 10, nburn)
  }
  p_value <- sum(random_delta > SYMdeltaA) / length(random_delta)
  smooth_p_value <- (sum(random_delta > SYMdeltaA) + 1) / (length(random_delta) + 1)

  # capture boxplot
  dev.new()
  boxplot(random_delta, main = titlestring)
  abline(h = SYMdeltaA, col = "red")
  boxplot_recorded <- recordPlot()
  dev.off()

  # capture prior plot
  dev.new()
  x <- seq(0, 10, length.out = 1000)
  plot(x, dgamma(x, shape = alpha, rate = beta), type = "l", main = "Prior distribution on delta")
  abline(v = SYMdeltaA, col = "red")
  prior_recorded <- recordPlot()
  dev.off()

  print("P-value")
  print(p_value)
  print("Smoothed p-value")
  print(smooth_p_value)

  # Return results
  list(deltaA = SYMdeltaA, p_value = p_value, smooth_p_value = smooth_p_value,  boxplot = boxplot_recorded, prior_plot = prior_recorded)
}

```


```{r}
set.seed(22)
h_results <- delta_plot_p(h_factor, htree, "Homelessness delta distribution")
replayPlot(h_results$boxplot)
```

```{r}
replayPlot(h_results$prior_plot)
```

```{r}
set.seed(22)
i_results <- delta_plot_p(i_factor, itree, "Illicit_ever delta distribution")
replayPlot(i_results$boxplot)
```

```{r}
replayPlot(i_results$prior_plot)
```

```{r}
##### conclusions:
#    for both variables, the deltaA estimate falls towards the tails of the (0.1, 0.05) prior distribution but not in the extreme tail. Much closer to  of (1, 1) distribution.
#    many samples dropped from both sets of test samples due to invalid trait value - started with 266; ended with 194 for homelessness, 183 for illicit_ever
#    distribution/proportions of trait values in both test sets were balanced and not skewed/biased
#    Both yielded significant results. That significance, as well as the observed deltaA value, remained stable and robust to the following tests:
         # different tree inputs with slightly altered sample composition - addition of Genbank relatives (ex tried Matt's original ML input for time tree, my own ML input for time tree)
         # Adjusting parameters in response to flat likelihood surface warning - gave chain longer to explore with higher iterations, permuations, and kept more effective samples with higher scaled burnin
         # encoding "No" and "Unknown" as "No/Unknown" like in JUNIPER - larger sample number, dropping only empty/nan
         # changing prior distribution specs: I used (0.1, 0.05) but also got significant results with (1, 1) and (2, 2)
#    the illicit_ever vector/subset sometimes (not always) generated warnings about flat likelihood surface for transition rate matrix, which means ape's ace() struggled to find a well-defined ML estimate 
         # The data didn't contain enough info to estimate certain rates confidently. This can result from many factors. 
         # the illicit_ever set is a bit more skewed (65/35) than the homelessness set. Maybe there aren't enough transitions away from dominant state ("Yes") to estimate the rate
         # tree itself could be unbalanced, pectinate, or just oddly clustered so that it doesn't have much internal node information. 
         # Not pectinate per Sackin. Visualize to see clustering - I do think that illicit_ever has some odd clumping, with relatively few "No" values

# # are our trees pectinate?
# # Check how many internal nodes have exactly one tip child - a fully pectinate tree will have Ntip-1 such nodes
# # Note: some degree of pectination is normal with large datasets
# # Sackin index - sum of all tip depths

# ## only one tree for encoded data
# node_depths <- node.depth(htree, method = 1)
# sackin <- sum(node_depths[1:Ntip(htree)])
# print(sackin) # [1] 194
# n <- Ntip(htree) # Normalize
# sackin_normalized <- sackin / (n * (n-1) / 2)
# print(" sackin")
# print(sackin_normalized)  # closer to 1 = more pectinate

# # ## illicit ever
# # node_depths <- node.depth(itree, method = 1)
# # sackin <- sum(node_depths[1:Ntip(itree)])
# # print(sackin) # [1] 183
# # n <- Ntip(itree) # Normalize
# # sackin_normalized <- sackin / (n * (n-1) / 2)
# # print("Illicit ever sackin")
# # print(sackin_normalized)  # closer to 1 = more pectinate


# # visualize to see distribution
# # Create color mapping 
# tip_colors <- hsubs$homelessness[match(htree$tip.label, hsubs$MASPHL_Sequencing_ID)]
# color_map <- ifelse(tip_colors == "Yes", "#E8A838",
#              ifelse(tip_colors == "No/Unknown", "#7B7FC4", NA))
# hsubs <- hsubs %>%
#   mutate(label = MASPHL_Sequencing_ID, .before = 1)
# hplot <- ggtree(htree) %<+% hsubs +
#   geom_tippoint(aes(color = homelessness), size = 3, shape = 16) +
#   scale_color_manual(values = c("Yes" = "#E8A838", "No/Unknown" = "#7B7FC4"),
#                      na.value = "grey80",
#                      breaks = c("Yes", "No/Unknown")) +
#   theme(legend.position = "right")

# plot(hplot)
```

```{r}
# # itree
# tip_colors <- isubs$illicit_ever[match(itree$tip.label, isubs$MASPHL_Sequencing_ID)]
# color_map <- ifelse(tip_colors == "Yes", "#E8A0B4",
#              ifelse(tip_colors == "No/Unknown", "#7B9EC4", NA))
# isubs <- isubs %>%
#   mutate(label = MASPHL_Sequencing_ID, .before = 1)
# iplot <- ggtree(itree) %<+% isubs +
#   geom_tippoint(aes(color = illicit_ever), size = 3, shape = 16) +
#   scale_color_manual(values = c("Yes" = "#E8A0B4", "No/Unknown" = "#7B9EC4"),
#                      na.value = "grey80",
#                      breaks = c("Yes", "No/Unknown")) +
#   theme(legend.position = "right")

# plot(iplot)
```

```{r}
##########################################
# Write summary
cat("=== TREE SUMMARY ===", "\n")
cat("Original tree tip count:", length(uditree$tip.label), "\n")
cat("Homelessness subset tree tip count:", length(htree$tip.label), "\n")
cat("Illicit ever subset tree tip count:", length(itree$tip.label), "\n")

cat("\n=== TRAIT VALUES ALLOWED ===", "\n")
cat("htrait unique values:", paste(unique(htrait), collapse = ", "), "\n")
cat("itrait unique values:", paste(unique(itrait), collapse = ", "), "\n")

cat("\n=== FACTOR BALANCE (table()) ===", "\n")
cat("h_factor:", "\n")
print(table(h_factor))
cat("i_factor:", "\n")
print(table(i_factor))

```

```{r}
cat("\n=== PARAMETERS ===")
cat("niter:", niter, "\n")
cat("nburn:", nburn, "\n")
cat("nperm:", nperm, "\n")
cat("alpha:", alpha, "\n")
cat("beta:", beta, "\n")

```

```{r}
cat("\n=== HOMELESSNESS DELTA RESULTS ===", "\n")
cat("deltaA:", h_results$deltaA, "\n")
cat("p-value:", h_results$p_value, "\n")
cat("smoothed p-value:", h_results$smooth_p_value, "\n")

```

```{r}

cat("\n=== ILLICIT EVER DELTA RESULTS ===", "\n")
cat("deltaA:", i_results$deltaA, "\n")
cat("p-value:", i_results$p_value, "\n")
cat("smoothed p-value:", i_results$smooth_p_value, "\n")

```

## NOTE: why did I initially get some 0 p-values with a smaller/more stringently filtered sample set?
Potentially have poor or noisy phylogenetic signal to begin with — note that deltaA ~0.56 is moderate
p value resolution is 0.01 so p=0 means p<0.01 - that is, none of the 100 random permutations exceeded the observed deltaA
suggests non-random clustering - need to test parameters though to see if deltaA stays stable
One option: change 100 --> 1000 permutations
     Also change iterations from 10000 to 50000 - flat likelihood surface means that the chain needs longer to explore properly; 10,000 iterations, keeping every 10th, gives you only 1,000 effective samples before burn-in, not good for a flat-gradient case
     scale burn-in proportionally to iterations - 500 discarded iterates fits better

## NOTE: If you get warnings like:
 Warning messages:
  1: In .getSEs(out.nlm) :
   The likelihood gradient seems flat in at least one dimension (gradient null):
  cannot compute the standard-errors of the transition rates.

These appear if the likelihood surface for the transition rate matrix is flat or nearly flat in at least one dimension, i.e., ape::ace() struggled to find a 
well-defined maximum likelihood estimate for some transition rates, which happens when the data don't contain enough information to estimate certain rates confidently.
Some potential reasons:
If one state is rare or clustered in a way that gives very few observed transitions between states, there's little information to estimate the rate for that transition. The optimizer finds a ridge rather than a clean peak.
A trait with many categories is more likely to have some rate parameters that are poorly informed by the data - more states = more rates to estimate
If your trait has heavily skewed or imbalanced states, ex 90% of tips are in one state, then transitions away from the dominant state are rare and their rates become hard to estimate.
If your tree is unbalanced or pectinate tree then it has less internal node information for some clades, which can make certain transitions unobservable.

As the README says, the standard errors of those rate parameters aren't actually used by δ, as it only uses the resulting node probability vectors. 
So a trait that triggers the warning isn't necessarily giving you a worse δ estimate, it just means the underlying rate model was harder to fit for that particular trait/tree combination.

```{r}
############################################################################
# # look at missingness for UNENCODED data - does missingness in one variable depend on any other covariates? investigate whether data is missing at random
# # note: this treats "Unknown" values the same as NA or empty string
# all = c("homelessness", "illicit_ever")
# explanatory = all[1]
# dependent = all[2]

# uesubs <- meta[match(tipnames, meta$MASPHL_Sequencing_ID ),]
# mdcopy=as.data.frame(uesubs[,c("MASPHL_Sequencing_ID","homelessness", "illicit_ever", "gender","county_of_residence")])
# mdcopy$homelessness <- gsub("Unknown", "", mdcopy$homelessness)
# mdcopy$illicit_ever <- gsub("Unknown", "", mdcopy$illicit_ever)
# mdcopy$gender <- gsub("Unknown", "", mdcopy$gender)
# mdcopy$county_of_residence <- gsub("Unknown", "", mdcopy$county_of_residence)

# mdcopy <- mdcopy %>%
#    mutate(across(everything(), ~ na_if(., "")))

# #https://finalfit.org/articles/missing.html
# mdcopy %>% 
#   summary_factorlist("homelessness",c("illicit_ever"), na_include=TRUE, p=TRUE)
# # Note: dependent includes missing data. These are dropped.
# #         label levels        No       Yes      p
# #  illicit_ever     No 52 (52.0)  9 (12.3) <0.001
# #                  Yes 48 (48.0) 64 (87.7)  
# # # position = stack is for counts; position = fill is for proportions; stack - counts by default
# # mdcopy %>%
# #   missing_pairs(dependent, explanatory, use_labels=TRUE, showXAxisPlotLabels=TRUE,showYAxisPlotLabels=TRUE)

# #######################################################################################
# # ## More in-depth examination of original four variables of interest
# mdcopy %>% 
#   summary_factorlist("gender", c("homelessness","illicit_ever"), na_include=TRUE, p=TRUE)
# # Note: dependent includes missing data. These are dropped.
# #         label levels        No       Yes      p
# #  illicit_ever     No 52 (52.0)  9 (12.3) <0.001
# #                  Yes 48 (48.0) 64 (87.7)     

# mdcopy %>% 
#   summary_factorlist("illicit_ever", c("homelessness","gender"), na_include=TRUE, p=TRUE)
# # Note: dependent includes missing data. These are dropped.
# #         label levels        No       Yes      p
# #  homelessness     No 52 (85.2) 48 (42.9) <0.001
# #                  Yes  9 (14.8) 64 (57.1)       
# #        gender Female 16 (25.4) 48 (40.0)  0.071
# #                 Male 47 (74.6) 72 (60.0)    

  
# all = c("homelessness", "illicit_ever", "gender", "county_of_residence")
# explanatory = all[2:4]
# dependent = all[1]
# # proportions
# mdcopy %>%
#   missing_pairs(dependent, explanatory, use_labels=TRUE, position = "fill", showXAxisPlotLabels=TRUE,showYAxisPlotLabels=TRUE)

# #ff_plot() will automatically determine the dependent variable type (binary, survival object, or continuous), and produce the appropriate output (odds ratio, hazard ratio, or coefficient plot).
# mdcopy %>%
#   ff_glimpse(dependent, explanatory)
# # $Continuous
# # data frame with 0 columns and 262 rows

# # $Categorical
# #                                   label var_type   n missing_n missing_percent levels_n levels levels_count levels_percent
# # homelessness               homelessness    <chr> 194        68            26.0        2      -            -              -
# # illicit_ever               illicit_ever    <chr> 183        79            30.2        2      -            -              -
# # gender                           gender    <chr> 213        49            18.7        2      -            -              -
# # county_of_residence county_of_residence    <chr> 212        50            19.1       11      -            -              -

# mdcopy %>%
#   missing_plot(dependent,explanatory)

# # #wraps mice::md.pattern using finalfit grammar.
# # # https://amices.org/mice/reference/md.pattern.html
# # # each row corresponds to a missing data pattern: 1=observed, 0=missing.
# # # The output tells us that 173 samples are complete, 21 samples miss only the illicit_ever measurement, 10 miss only homelessness, etc
# # # some are missing all four of them

# # # 49 rows are missing all four of them (complete missingness)
# mdcopy %>% 
#    missing_pattern(dependent, explanatory)
# #     gender county_of_residence homelessness illicit_ever    
# # 173      1                   1            1            1   0
# # 21       1                   1            1            0   1
# # 10       1                   1            0            1   1
# # 8        1                   1            0            0   2
# # 1        1                   0            0            0   3
# # 49       0                   0            0            0   4
# #         49                  50           68           79 246
# #####################################

```