---
title: "Figure 1"
execute:
  include: false
---

**Figure 1. Total confirmed and sequenced HAV cases from late 2018 through early 2024.** 

**A)** Top: Total confirmed cases in MA, shaded by whether cases were successfully sequenced in this study (left axis) and HAV vaccines administered in MA per 100,000 people (right axis). Bottom: outbreak-associated cases in NH, shaded by whether cases were successfully sequenced in this study. 

**B)** HAV vaccines administered by vaccine provider type. 

**C)** Total outbreak-associated cases by MA town. 

**D)** Total sequenced cases by MA town, including both outbreak-associated and non-outbreak-associated cases.  

```{r}
rm(list = ls())
pkglist = c("tinytex", "ggplot2", "ggpattern", "ggpubr", "vegan", "stringr", "splitstackshape")
lapply(pkglist, require, character.only = TRUE)
```

```{r}
OUTPUT_DIR=Sys.getenv("OUTPUT_DIR")
COUNTY_SHAPE=Sys.getenv("COUNTY_SHAPE")
CLEANING_SCRIPT=Sys.getenv("CLEANING_SCRIPT")
source(CLEANING_SCRIPT)
```

```{r}
#################################
##### Load and prepare data #####
#################################
# we were going to include CT and NH cases initially, but have since to at least hold off on that for now. However, Lydia still needs the start of the NH shape for her plot, so only excluding CT with this subset
metadata_town_clean <- subset(metadata_town_clean, cityResidence!="Connecticut")

##### Shapefiles
# town shapes
town_shape <- read_sf(TOWN_SHAPE)
# the TOWN variable gives all capitals for all town names, so just changing it so only the first letter of every word is capitalized.
town_shape$town <- stri_trans_totitle(town_shape$TOWN)
# county shapes
county_shape <- read_sf(COUNTY_SHAPE)
# both datasets have a column called "geometry" which describes the shape of the object in that row (either town or county). So I am changing the name of the "geometry" column in the county_shape dataframe to "countyGeo" so that I can have county and town shape variables in one dataset in later merge steps.
names(county_shape)[names(county_shape)=="geometry"] <- "countyGeo"

##### Processing cases by town
# Merge in town shape to cases_town_all. I went to do this in the cleaning program, but excel can't handle the export of geometry scf data (kept giving null values), so had to create the desired shape file and then load/merge it in here.
cases_town_all <- merge(metadata_town_clean,
                        town_shape,
                        by.x="cityResidence",
                        by.y="town",
                        all.x=TRUE)

# merge in county shape to cases_town_all.
cases_town_all <- merge(cases_town_all,
                        county_shape,
                        by.all="county",
                        all.x=TRUE)

# we have NH data; currently, that row is identified using the cityResidence variable (cityResidence=NH for those out-of-state-cases). Adding NH as a county if not an MA town
cases_town_all$county <- ifelse(cases_town_all$cityResidence != "New Hampshire", cases_town_all$county, "New Hampshire")

# add a county-case density variable for proper coloring of the choropleth that will match with the phylogenetic tree
#cases_town_all$townCaseGrp6 <- paste0(cases_town_all$sequencedCtGrp6, "-", cases_town_all$county)
cases_town_all$iiiaTownGrp <- paste0(cases_town_all$iiiaCtGrp3, "-", cases_town_all$county)

# Factoring variables for plotting so count/rate groups aren't lost in the legend labels if there are no observations of a certain count/rate group.
cases_town_all$overallCtGrp6 <- factor(cases_town_all$overallCtGrp6,
                                       levels=c("0",
                                                "1-5",
                                                "6-10",
                                                "11-30",
                                                "31-60",
                                                ">60"))

cases_town_all$sequencedCtGrp6 <- factor(cases_town_all$sequencedCtGrp6,
                                         levels=c("0",
                                                  "1-5",
                                                  "6-10",
                                                  "11-30",
                                                  "31-60",
                                                  ">60"))

cases_town_all$outbreakCtGrp6 <- factor(cases_town_all$outbreakCtGrp6,
                                        levels=c("0",
                                                 "1-5",
                                                 "6-10",
                                                 "11-30",
                                                 "31-60",
                                                 ">60"))

cases_town_all$outbreak1OverallCtGrp6 <- factor(cases_town_all$outbreak1OverallCtGrp6,
                                                levels=c("0",
                                                         "1-5",
                                                         "6-10",
                                                         "11-30",
                                                         "31-60",
                                                         ">60"))

cases_town_all$outbreak1SequencedCtGrp6 <- factor(cases_town_all$outbreak1SequencedCtGrp6,
                                                  levels=c("0",
                                                           "1-5",
                                                           "6-10",
                                                           "11-30",
                                                           "31-60",
                                                           ">60"))

cases_town_all$outbreak1OutbreakCtGrp6 <- factor(cases_town_all$outbreak1OutbreakCtGrp6,
                                                 levels=c("0",
                                                          "1-5",
                                                          "6-10",
                                                          "11-30",
                                                          "31-60",
                                                          ">60"))

cases_town_all$outbreak2OverallCtGrp3 <- factor(cases_town_all$outbreak2OverallCtGrp3,
                                                levels=c("0",
                                                         "1-5",
                                                         ">5"))

cases_town_all$outbreak2SequencedCtGrp3 <- factor(cases_town_all$outbreak2SequencedCtGrp3,
                                                  levels=c("0",
                                                           "1-5",
                                                           ">5"))

cases_town_all$outbreak2OutbreakCtGrp3 <- factor(cases_town_all$outbreak2OutbreakCtGrp3,
                                                 levels=c("0",
                                                          "1-5",
                                                          ">5"))

cases_town_all$outbreak1OverallRtGrp6 <- factor(cases_town_all$outbreak1OverallRtGrp6,
                                                levels=c("0",
                                                         ">0-5",
                                                         ">5-10",
                                                         ">10-30",
                                                         ">30-60",
                                                         ">60"))

cases_town_all$outbreak1SequencedRtGrp6 <- factor(cases_town_all$outbreak1SequencedRtGrp6,
                                                  levels=c("0",
                                                           ">0-5",
                                                           ">5-10",
                                                           ">10-30",
                                                           ">30-60",
                                                           ">60"))

cases_town_all$outbreak2OverallRtGrp3 <- factor(cases_town_all$outbreak2OverallRtGrp3,
                                                levels=c("0",
                                                         ">0-5",
                                                         ">5"))

cases_town_all$outbreak2SequencedRtGrp3 <- factor(cases_town_all$outbreak2SequencedRtGrp3,
                                                  levels=c("0",
                                                           ">0-5",
                                                           ">5"))

# # Check the actual min/max of your date column
# range(cases_week_all$endDt, na.rm=TRUE)
# # [1] "1970-01-01" "2024-11-23"

# # This is the Unix epoch date, so R didn't coerce this value to a date correctly. 
# # Find the offending rows
# cases_week_all %>% filter(endDt == as.Date("1970-01-01"))
# # wkYr=0, endDt = "2024-11-23"
# head(cases_week_all)

# cases_week_all %>% filter(endDt == as.Date("1970-01-01"))

# # Check if any dates fall outside the limits you set
# cases_week_all %>% 
#   filter(endDt < min(cases_week_all$endDt) | endDt > max(cases_week_all$endDt))

# # Also check for NAs
# sum(is.na(cases_week_all$endDt))
```

```{r}
#########################################################
##### Figure 1 - Outbreak cases and sequenced cases #####
#########################################################
##### epi outbreak curve for MA
fig1_a <- ggplot() +
  ##### epi curve
  # non-outbreak cases stacked on top the sequenced and non-sequenced outbreak cases
  geom_col(data=cases_week_all,
           aes(x=endDt,
               y=outNoseqYes+outNoseqNo+outYesseqNo+outYesseqYes,
               fill="Non-outbreak")) +
  # non-sequenced outbreak cases on top of the sequenced cases
  geom_col(data=cases_week_all,
           aes(x=endDt,
               y=outYesseqNo+outYesseqYes,
               fill="Outbreak (not sequenced)")) +
  # sequenced cases as the bottom layer
  geom_col(data=cases_week_all,
           aes(x=endDt,
               y=outYesseqYes,
               fill="Sequenced outbreak")) +
  ##### outbreak markers
  # line for 2018-2020 outbreak notice
  geom_segment(aes(x=as.Date("2018-08-22"),
                   xend=as.Date("2018-08-22"),
                   y=0,
                   yend=29),
               linetype=3,
               color="black") +
  # line for 2023-2024 outbreak notice
  geom_segment(aes(x=as.Date("2023-12-04"),
                   xend=as.Date("2023-12-04"),
                   y=0,
                   yend=29),
               linetype=3,
               color="black") +
  # shading the COVID emergency response period
  annotate("rect",
           xmin=as.Date("2020-03-10"),
           xmax=as.Date("2021-06-15"),
           ymin=0,
           ymax=29,
           alpha=0.1,
           fill="#808080") +
  geom_errorbarh(aes(xmin = as.Date("2020-03-10"), xmax = as.Date("2021-06-15"),
                 y=28.75),
                 colour = "#4D4D4D",
                 size=0.4) +
  # line for rolling 4-week average of vaccines administered per 100,000 residents. I have to subset the NA cases out (the first few rows in the rolling vaccination rate calculation) so I don't get an annoying error.
  geom_line(data=subset(cases_week_all, !is.na(rollingVaxPer100000)),
            aes(x=endDt,
                y=rollingVaxPer100000,
                color="Vaccinations per\n100,000 people")) +
  ##### plot labels
  labs(x="Week end date",
       y="MA case counts",
       title="a") +
  ##### plot color choices
  scale_fill_manual(name="Case type",
                    breaks=c("Sequenced outbreak", "Outbreak (not sequenced)", "Non-outbreak"),
                    values=c("Sequenced outbreak"="#4719E4",
                             "Outbreak (not sequenced)"="#9E8AE0",
                             "Non-outbreak"="lightgray"),
                    drop=FALSE) +
  ##### axis settings
  scale_y_continuous(expand = c(0, 0),
                     breaks=c(0,5,10,15,20,25),
                     sec.axis=sec_axis(~., 
                                       name="Vaccinations per\n100,000 people", 
                                       breaks=c(0,5,10,15,20,25))) +
  scale_x_date(expand = c(0, 0),
               # I wanted to manually choose the breaks as yearly on January, so with my knowledge doing it manually was the way to go
               breaks=c(as.Date("2019-01-01"), 
                        as.Date("2020-01-01"), 
                        as.Date("2021-01-01"), 
                        as.Date("2022-01-01"), 
                        as.Date("2023-01-01"), 
                        as.Date("2024-01-01")),
               date_labels="%b %Y",
               # creating a fake secondary x-axis for the 2018-2020 outbreak response label. In the theme(), I have erased all aspects of this axis, but have made a custom label. This allowed me to label the plot outside of the plotting area.
               sec.axis = sec_axis(~.,
                                   # now adding the 2018-2020 outbreak label
                                   breaks=c(as.Date("2018-05-15"), as.Date("2020-03-05"), as.Date("2023-09-07")),
                                   labels=c("Statewide outbreak\nresponse initiated\nAug 2018", "MA COVID-19 \nstate of emergency \ndeclaration", "Statewide outbreak\nresponse initiated\nDec 2023"))) +
  ## using this to expand the plot area a bit to allow me to label a bit better without labels being cut off. I tried to do this to label above the plot, but it was adding length to the x-axis which was not desirable.
  coord_cartesian(clip="off", expand=FALSE, ylim=c(0,28)) +
  ##### plot theme
  theme_bw() +
  theme(panel.grid = element_blank(),
        text = element_text(family = "sans"),
        strip.background = element_blank(),
        axis.text=element_text(size=9),
        axis.title=element_text(size=9),
        plot.title=element_text(size=12, hjust=-0.09),
        legend.text=element_text(size=9),
        legend.title=element_text(size=9),
        strip.text = element_text(size = 9),
        panel.border = element_blank(),
        axis.line = element_line(colour = "black"),
        # get rid of top x-axis that we created for the 2018-2020 outbreak response.
        axis.ticks.x.top = element_blank(),
        axis.text.x.top = element_text(margin = margin(b = 5), hjust=0, size=9, family="sans", color="black"),
        axis.line.x.top = element_blank(),
        plot.margin = unit(c(0,-0.2,0,0.4), "cm"),
        legend.key.size = unit(0.2, "cm"),
        panel.background = element_rect(fill = "transparent",
                                        colour = NA_character_),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        plot.background = element_rect(fill = "transparent",
                                       colour = NA_character_),
        legend.background = element_rect(fill = "transparent"),
        legend.box.background = element_rect(fill = "transparent", colour="transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.spacing.y = unit(-0.5, "cm")) +
  # color for vaccination rate
    scale_color_manual(name="",
                       breaks="Vaccinations per\n100,000 people",
                       values=c('Vaccinations per\n100,000 people'='#818181'))
plot(fig1_a)
```

```{r}
##### epi outbreak curve for NH
fig1_a2 <- ggplot() +
  ##### epi curve
  # non-outbreak cases stacked on top the sequenced and non-sequenced outbreak cases
  geom_col(data=cases_week_all,
           aes(x=endDt,
               y=nhTotWkCaseCt,
               fill="Outbreak (not sequenced)")) +
  geom_col(data=cases_week_all,
           aes(x=endDt,
               y=nhSeqWkCaseCt,
               fill="Sequenced outbreak")) +
  ##### plot labels
  labs(x="Week end date",
       y="NH case counts") +
  ##### axis settings
  scale_y_continuous(expand = c(0, 0),
                     breaks=c(0,5,10,15,20,25)) +
  # using this to expand the plot area a bit to allow me to label a bit better without labels being cut off. I tried to do this to label above the plot, but it was adding length to the x-axis which was not desirable.
  coord_cartesian(clip="off", expand=FALSE, ylim=c(0,28)) +
  ##### plot color choices
  scale_fill_manual(name="Case type",
                    breaks=c("Sequenced outbreak", "Outbreak (not sequenced)"),
                    values=c("Sequenced outbreak"="black",
                             "Outbreak (not sequenced)"="darkgray"),
                    drop=FALSE) +
  ##### plot theme
  theme_bw() +
  theme(panel.grid = element_blank(),
        text = element_text(family = "sans"),
        strip.background = element_blank(),
        axis.text=element_text(size=9),
        plot.title=element_text(size=12, hjust=-0.055),
        legend.text=element_text(size=9),
        legend.title=element_text(size=9),
        strip.text = element_text(size = 9),
        panel.border = element_blank(),
        axis.line = element_line(colour = "black"),
        # get rid of top x-axis that we created for the 2018-2020 outbreak response.
        axis.ticks.x.top = element_blank(),
        axis.text.x.top = element_text(margin = margin(b = 5), hjust=0, size=9, family="sans", color="black"),
        axis.line.x.top = element_blank(),
        legend.key.size = unit(0.2, "cm"),
        panel.background = element_rect(fill = "transparent",
                                        colour = NA_character_),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        plot.background = element_rect(fill = "transparent",
                                       colour = NA_character_),
        legend.background = element_rect(fill = "transparent"),
        legend.box.background = element_rect(fill = "transparent", colour="transparent"),
        legend.key = element_rect(fill = "transparent"))

plot(fig1_a2)
```

```{r}
# plotting vaccines administered by provider
fig1_b <- ggplot() +
##### data
# pediatric doses
  geom_col(data=subset(cases_week_all, !is.na(rollingWkImmCt_prov2Pediatric)),
            aes(x=endDt,
                y=rollingWkImmCt_prov2HealtCent+rollingWkImmCt_prov2CommHealthCent+rollingWkImmCt_prov2CommPharm+rollingWkImmCt_prov2Pediatric,
                fill="Pediatrician")) +
  # pharmacy doses
  geom_col(data=subset(cases_week_all, !is.na(rollingWkImmCt_prov2CommPharm)),
           aes(x=endDt,
               y=rollingWkImmCt_prov2HealtCent+rollingWkImmCt_prov2CommHealthCent+rollingWkImmCt_prov2CommPharm,
               fill="Commercial pharmacy")) +
  # community health center doses
  geom_col(data=subset(cases_week_all, !is.na(rollingWkImmCt_prov2CommHealthCent)),
           aes(x=endDt,
               y=rollingWkImmCt_prov2HealtCent+rollingWkImmCt_prov2CommHealthCent,
               fill="Community health center")) +
  # healthcare center doses
  geom_col(data=subset(cases_week_all, !is.na(rollingWkImmCt_prov2HealtCent)),
           aes(x=endDt,
               y=rollingWkImmCt_prov2HealtCent,
               fill="Healthcare center")) +
##### outbreak markers
# 2018-2020 outbreak marker
  geom_vline(aes(xintercept=as.Date("2018-08-22")),
             linetype=3,
             color="black") +
# 2023-2024 outbreak marker
  geom_vline(aes(xintercept=as.Date("2023-12-04")),
             linetype=3,
             color="black") +
# COIVD-19 outbreak marker
  annotate("rect",
           xmin=as.Date("2020-03-10"),
           xmax=as.Date("2021-06-15"),
           ymin=0,
           ymax=1350,
           alpha=0.1,
           fill="#808080") +
##### theme
  theme_bw() +
  theme(text = element_text(family = "sans"),
        axis.line.x=element_line(color="black"),
        axis.line.y=element_line(color="black"),
        axis.text=element_text(size=8),
        axis.title=element_text(size=9),
        legend.text=element_text(size=9),
        legend.title=element_text(size=9),
        plot.title=element_text(size=12, hjust=-0.09),
        panel.border = element_blank(),
        axis.line = element_line(colour = "black"),
        axis.ticks.x.top = element_blank(),
        axis.text.x.top = element_text(margin = margin(b = 5), hjust=0, size=9, family="sans"),
        axis.line.x.top = element_blank(),
        legend.key.size = unit(0.2, "cm"),
        plot.margin = unit(c(0,1.21,0,0.125), "cm"),
        panel.background = element_rect(fill = "transparent",
                                        colour = NA_character_),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        plot.background = element_rect(fill = "transparent",
                                       colour = NA_character_),
        legend.background = element_rect(fill = "transparent"),
        legend.box.background = element_rect(fill = "transparent", colour="transparent"),
        legend.key = element_rect(fill = "transparent")) +
##### axis settings
  scale_y_continuous(expand = c(0, 0), breaks=c(0,150,300,450,600,750,900, 1050, 1200, 1350)) +
  scale_x_date(expand = c(0, 0),
               breaks=c(as.Date("2019-01-01"), 
                        as.Date("2020-01-01"), 
                        as.Date("2021-01-01"), 
                        as.Date("2022-01-01"), 
                        as.Date("2023-01-01"), 
                        as.Date("2024-01-01")),
               date_labels="%b %Y") +
##### plot labels
  labs(x="Week end date",
       y="Vaccine count",
       fill="Vaccine provider type",
       title="b") +
  scale_fill_manual(breaks=c("Other",
                            "Pediatrician",
                            "Commercial pharmacy",
                             "Community health center",
                            "Healthcare center"),
                    values=c("Other"="#C3C3C3",
                             "Pediatrician"="#A8FFB2",
                             "Commercial pharmacy"="#4CE847",
                             "Community health center"="#149E24",
                             "Healthcare center"="#0D5816")) +
  coord_cartesian(clip = "off")

plot(fig1_b)
```

```{r}
##### choropleth of outbreak case count by town
fig1_c <- ggplot(data=subset(cases_town_all, cityResidence != "New Hampshire")) +
  ##### data and town shape
  geom_sf(aes(geometry=geometry, fill=outbreakCtGrp6),
          show.legend=TRUE,
          color=NA) +
  ##### theme
  theme_void() +
  theme(text = element_text(family = "sans"),
        legend.text=element_text(size=9),
        legend.title=element_text(size=9),
        plot.title=element_text(size=12),
        legend.key.size = unit(0.2, "cm")) +
  ##### removing odd lines around legend color boxes
  guides(fill=guide_legend(override.aes=list(linetype=c(0,0,0,0,0,0)))) +
  ##### plot color choices
  scale_fill_manual(breaks=c("0",
                             "1-5",
                             "6-10",
                             "11-30",
                             "31-60",
                             ">60"),
                    values=c("0"="lightgray",
                             "1-5"="#f0ff00",
                             "6-10"="#ffcd32",
                             "11-30"="#ff9700",
                             "31-60"="#ff5d13",
                             ">60"="#ff0000"),
                    name="MA outbreak case count",
                    drop=FALSE) +
  ##### plot labels
  labs(title="c")

plot(fig1_c)
```

```{r}
##### choropleth of sequenced case count by town
fig1_d <- ggplot(data=subset(cases_town_all, cityResidence != "New Hampshire")) +
  ##### data and town shape
  geom_sf(aes(geometry=geometry, fill=sequencedCtGrp6),
          show.legend=TRUE,
          color=NA) +
  ##### theme
  theme_void() +
  theme(text = element_text(family = "sans"),
        legend.text=element_text(size=9),
        legend.title=element_text(size=9),
        plot.title=element_text(size=12),
        legend.key.size = unit(0.2, "cm")) +
  ##### removing odd lines around legend color boxes
  guides(fill=guide_legend(override.aes=list(linetype=c(0,0,0,0,0,0)))) +
  ##### plot color choices
  scale_fill_manual(breaks=c("0",
                             "1-5",
                             "6-10",
                             "11-30",
                             "31-60",
                             ">60"),
                    values=c("0"="lightgray",
                             "1-5"="#f0ff00",
                             "6-10"="#ffcd32",
                             "11-30"="#ff9700",
                             "31-60"="#ff5d13",
                             ">60"="#ff0000"),
                    name="MA sequenced case count",
                    drop=FALSE) +
  ##### plot labels
  labs(title="d")

plot(fig1_d)
```

```{r}
##### now creating the plot panel of the epi curve (sequenced vs non-sequenced), total cases by town, and total sequenced cases by town.
fig1 <- ggarrange(fig1_a,
                  fig1_a2 +
                    theme(plot.margin = unit(c(0,1.06,0,0.4), "cm"),
                          axis.title=element_text(size=9)) +
                    scale_x_date(expand = c(0, 0),
                                 # I wanted to manually choose the breaks as yearly on January, so with my knowledge doing it manually was the way to go
                                 breaks=c(as.Date("2019-01-01"), 
                                          as.Date("2020-01-01"), 
                                          as.Date("2021-01-01"), 
                                          as.Date("2022-01-01"), 
                                          as.Date("2023-01-01"), 
                                          as.Date("2024-01-01")),
                                 date_labels="%b %Y"),
                  fig1_b,
                  ggarrange(fig1_c +
                              theme(plot.margin = unit(c(-3,0,-2,0.125), "cm")),
                            fig1_d +
                              theme(plot.margin = unit(c(-3,0.125,-2,0), "cm")), 
                            ncol=2,
                            align="v"),
                  nrow=4,
                  heights=c(1.2,0.6,1,1))

plot(fig1)
```

```{r}
##### pdf of Figure 1
#fig1 %>%
ggsave(paste(OUTPUT_DIR, "/Fig1.pdf", sep="/"), fig1, height=7, width=8, units="in", dpi=300)

##### png of Figure 1
#fig1 %>%
ggsave(paste(OUTPUT_DIR, "Fig1.png", sep="/"), fig1, height=7, width=7, units="in", dpi=300, bg="transparent")

```