---
title: "Figure 3"
execute:
  echo: false
---

**Figure 3. Transmission reconstruction of clade IIIA cases using epidemiological investigation and genomic data.**

**A**) Transmission links identified through epidemiological investigation (MA only) or inferred by JUNIPER with ≥50% support (MA and NH). Boxes highlight instances in which transmission reconstruction suggests an alternative infection pathway involving a third sequenced case not identified by epidemiological linkage. Node colors indicate MA county of residence, as in **Fig 2B**. 

**B)** For MA cases only, distributions of single-nucleotide polymorphisms (SNPs) for defined sets of case pairs. SNP distances between 0 and 15 are shown (see **Fig S6** for full distribution). 

**C)** For MA cases only, mean probability of initiating at least one transmission (left) and mean number of transmissions initiated across JUNIPER iterations (right) among PWUD/non-PWUD (top) and PEH/non-PEH (bottom). Density curves sum to 1. 

```{r}
rm(list = ls())
pkglist = c("igraph", "dplyr", "ggplot2",#3A and #3B
                "ggplot2", "data.table", "tidyr", #3C
                "cli", "devtools", "lmPerm", ## had to manually delete cli.dll and install with install.packages to update to adequate version
                "coda", "stringr", "cowplot", "plyr", "readxl", "tibble") #"TransPhylo", 

lapply(pkglist, require, character.only = TRUE)

```

```{r}
JUNIPER_HELPER_SCRIPT=Sys.getenv("JUNIPER_HELPER_SCRIPT")
EPI_LINKAGE_SCRIPT=Sys.getenv("EPI_LINKAGE_SCRIPT")
OUTPUT_DIR=Sys.getenv("OUTPUT_DIR")
RENAMED_ALL_PAIRS_SNP_DIST=Sys.getenv("RENAMED_ALL_PAIRS_SNP_DIST")

source(JUNIPER_HELPER_SCRIPT)
source(EPI_LINKAGE_SCRIPT)
```

```{r}
# # 3A
# epi links here = all-vs-all QC-passing epi pairs derived from linkage groups
all_links <- combined
all_links$support <- round(all_links$support * 100)
epi_links <- subset(all_links, type == "epi" | type == "epi and juniper")
nrow(epi_links) # n = 31

county_colors <- c(
  "NH"                = "#808080",
  "CT"                = "#000000",
  "Berkshire County"  = "#680000",
  "Franklin County"   = "#a90000",
  "Hampshire County"  = "#de2300",
  "Hampden County"    = "#ea6800",
  "Worcester County"  = "#d1a100",
  "Middlesex County"  = "#93d200",
  "Essex County"      = "#b4ff06",
  "Suffolk County"    = "#2de000",
  "Norfolk County"    = "#00af9e",
  "Bristol County"    = "#0485e4",
  "Plymouth County"   = "#254bc8",
  "Barnstable County" = "#2c2071"
)

# open output pdf
output_file <- paste0(OUTPUT_DIR, "/Fig3A.pdf")
pdf(output_file, width = 10, height = 6)

# epi and juniper links
par(mar = c(0, 0, 0, 0))
g <- graph_from_data_frame(all_links, directed = TRUE) # directed=TRUE makes it treat edges as directional (arrows) rather than undirected connections (i.e., each edge goes from-->to)
V(g)$county <- metadata$county_of_residence[match(V(g)$name, metadata$MASPHL_Sequencing_ID)]
V(g)$color <- county_colors[V(g)$county]
V(g)$color[is.na(V(g)$color)] <- "lightgray"
V(g)$event_date <- metadata$event_date[match(V(g)$name, metadata$MASPHL_Sequencing_ID)]

clustering <- cluster_walktrap(g)
clusters <- clustering$membership

centroids <- sapply(unique(clusters), function(cluster_id)
{
  nodes_in_cluster <- which(clusters == cluster_id)
  subgraph <- induced_subgraph(g, nodes_in_cluster)
  centroid <- which.min(centralization.degree(subgraph)$centralization)
  return(nodes_in_cluster[centroid])
})

V(g)$label <- NA
for (centroid in centroids)
{
  cluster_nodes <- which(clusters == clusters[centroid])
  min_event_date <- min(V(g)$event_date[cluster_nodes])
  V(g)$label[centroid] <- as.character(min_event_date)
}

set.seed(123)
fixed_layout <- layout_with_fr(g)
plot(g,
     layout = fixed_layout,
     #layout = layout_with_fr(g), #force-directed layout: uses a stochastic algorithm internally (the Fruchterman-Reingold algorithm) - without a seed, node positions can vary on each run 
     edge.arrow.mode = ifelse(E(g)$type == "epi", "-", ">"),
     edge.arrow.size = 0.3,
     edge.color = ifelse(E(g)$type == "epi", "#bfbfbf", ifelse(E(g)$type == "epi and juniper", "#0487c8", "black")),
     edge.lty = 1,
     edge.width = 1.5,
     edge.label = E(g)$support,
     edge.label.family = "Helvetica",
     edge.label.cex = 0.3,
     edge.label.color = "black",
     edge.label.dist = 1,        # <-- distance from edge midpoint
     edge.label.degree = pi/2,   # <-- offset angle (90 degrees = perpendicular to edge)
     vertex.label = NA,
     vertex.size = 3,
     vertex.color = V(g)$color,
     vertex.frame.color = ifelse(V(g)$name %in% passed, "black", NA))

dev.off()

```

```{r}
# visualize in line
plot(g,
     layout = fixed_layout, #layout_with_fr(g),
     edge.arrow.mode = ifelse(E(g)$type == "epi", "-", ">"),
     edge.arrow.size = 0.3,
     edge.color = ifelse(E(g)$type == "epi", "#bfbfbf", # light gray
                  ifelse(E(g)$type == "epi and juniper", "#0487c8", # bright blue/cyan blue
                  "black")),
     edge.lty = 1,
     edge.width = 1.5,
     edge.label = E(g)$support,
     edge.label.family = "Helvetica",
     edge.label.cex = 0.3,
     edge.label.color = "black",
     edge.label.dist = 1,        # <-- distance from edge midpoint
     edge.label.degree = pi/2,   # <-- offset angle (90 degrees = perpendicular to edge)
     vertex.label = NA,
     vertex.size = 3,
     vertex.color = V(g)$color,
     vertex.frame.color = ifelse(V(g)$name %in% passed, "black", NA))
```

```{r}
####################################################
# subset all-vs-all SNP distances to the three subsets
all_pairs <- read.table(RENAMED_ALL_PAIRS_SNP_DIST, sep="\t", header=TRUE, stringsAsFactors=TRUE)
nrow(all_pairs)

# drop anything with HepA/USA/MA/0046/2018 from all_pairs
all_pairs <- all_pairs[all_pairs$sequence1 != "HepA/USA/MA/0046/2018" & 
                       all_pairs$sequence2 != "HepA/USA/MA/0046/2018", ]

metadata <- read.csv(METADATA_FILE)
# also drop HepA/USA/MA/0046/2018 from metadata
metadata <- metadata[metadata$MASPHL_Sequencing_ID != "HepA/USA/MA/0046/2018", ]

# Get IDs that satisfy all conditions simultaneously (from MA, passed QC, subgenotype IIIA)
valid_ids <- metadata$MASPHL_Sequencing_ID[
  metadata$State.of.Residence == "MA" &
  metadata$Sample_Passed_QC == "Yes" &
  metadata$Subgenotype == "IIIA"]

print(length(valid_ids))

# Filter pairs where BOTH samples meet all criteria
all_pairs <- all_pairs[all_pairs$sequence1 %in% valid_ids & 
                       all_pairs$sequence2 %in% valid_ids, ]
nrow(all_pairs) 

colnames(all_pairs) <- c("sample1", "sample2", "SNP_distance")

# first, epi links only (31)
epi_pairs <- combined[combined$type %in% c("epi", "epi and juniper"), ]
nrow(epi_pairs)
epi_subset <- all_pairs[mapply(function(s1, s2) {any((epi_pairs$from == s1 & epi_pairs$to == s2) | (epi_pairs$from == s2 & epi_pairs$to == s1))}, all_pairs$sample1, all_pairs$sample2),] # n=31

# second, juniper links only
juniper_pairs <- combined[combined$type %in% c("juniper", "epi and juniper"), ]
nrow(juniper_pairs)# n = 78
juniper_subset <- all_pairs[mapply(function(s1, s2) {any((juniper_pairs$from == s1 & juniper_pairs$to == s2) | (juniper_pairs$from == s2 & juniper_pairs$to == s1))},  all_pairs$sample1, all_pairs$sample2),] # n= 78

#######################################
# ## third, all pairs from same county where both known symptom onset dates within 0-57 days
# # (also consider filtering by sequenced and passed QC samples)
meta_filtered <- metadata %>%
  filter(State.of.Residence == "MA",
         Sample_Passed_QC == "Yes",
         Subgenotype=="IIIA",
         !is.na(symptom_onset_date))

meta_filtered <- meta_filtered %>%
  filter(trimws(symptom_onset_date) != "")

print(nrow(meta_filtered))
# Get the valid IDs
valid_ids <- meta_filtered$MASPHL_Sequencing_ID

# Subset snps to pairs where both samples are in valid_ids
# Make sure dates are Date type, then filter to 0-57 day difference
dates_within_range <- all_pairs %>%
  filter(sample1 %in% valid_ids, sample2 %in% valid_ids) %>%
  left_join(meta_filtered %>% select(MASPHL_Sequencing_ID, onset1 = symptom_onset_date, county1 = county_of_residence),
            by = c("sample1" = "MASPHL_Sequencing_ID")) %>%
  left_join(meta_filtered %>% select(MASPHL_Sequencing_ID, onset2 = symptom_onset_date, county2 = county_of_residence),
            by = c("sample2" = "MASPHL_Sequencing_ID")) %>%
  mutate(onset1 = as.Date(onset1, format = "%m/%d/%Y"),
         onset2 = as.Date(onset2, format = "%m/%d/%Y"),
         date_diff = abs(as.numeric(difftime(onset1, onset2, units = "days")))) %>%
  filter(!is.na(onset1), !is.na(onset2),
         date_diff <= 57,
         county1 == county2)

print(nrow(dates_within_range))
print(nrow(all_pairs))

length(unique(c(dates_within_range$sample1, dates_within_range$sample2)))

```

```{r}
epi_links_label <- "Epidemiological links"
juniper_links_label <- "JUNIPER links with ≥50% support"
dates_within_range_label <- "All pairs from same county with known symptom onset dates within 0-57 days"
all_pairs_label <- "All pairs"

# combine into one table
epi_subset <- epi_subset %>% mutate(pair_type = epi_links_label)
juniper_subset <- juniper_subset %>% mutate(pair_type = juniper_links_label)
dates_within_range <- dates_within_range %>% mutate(pair_type = dates_within_range_label)
all_pairs <- all_pairs %>% mutate(pair_type = all_pairs_label)

combined_table <- bind_rows(epi_subset, juniper_subset, dates_within_range, all_pairs)
combined_table$pair_type <- factor(combined_table$pair_type, levels=c(epi_links_label, juniper_links_label, dates_within_range_label, all_pairs_label))

# plot
snp_distances_plot <- ggplot(combined_table, aes(x = SNP_distance)) +
  geom_histogram(binwidth = 1, aes(color = pair_type, fill = pair_type)) +
  facet_wrap(pair_type ~., scales="free_y", ncol=1, labeller = label_wrap_gen(width = 50)) +
  # scale_y_log10() +
  xlab("SNP distance") + ylab("Count") +
  scale_color_manual(values = setNames(
    c("black", "black", "black", "black"),
    c(epi_links_label, juniper_links_label, dates_within_range_label, all_pairs_label))) +
  scale_fill_manual(values = setNames(
    c("black", "black", "black", "black"),
    c(epi_links_label, juniper_links_label, dates_within_range_label, all_pairs_label))) +
  theme_bw() +
  theme(legend.position = "none",
        axis.line = element_blank(),
        panel.grid = element_blank(),
        strip.background = element_rect(fill = "white", color = "white"),
        strip.text = element_text(hjust = 0))

snp_distances_plot_zoomed <- snp_distances_plot + xlim(-1, 15)

plot(snp_distances_plot_zoomed)

width <- 4
height <- 5.5
ggsave(filename=paste0(OUTPUT_DIR, "/Fig3B_zoomed.pdf"), snp_distances_plot_zoomed, height=height, width=width, units="in", dpi=300)
ggsave(filename=paste0(OUTPUT_DIR, "/Fig3B_zoomed.png"), snp_distances_plot_zoomed, height=height, width=width, units="in", dpi=300)
```

```{r setup, include=FALSE}
# 3C
#Taylor's metadata filtering code from hav_25mar25_analysis.R: "about the metadata file then, I think it was just filtered for those sequences that were 95% complete and had passed a previously defined QC"
# he ended up with 305 in his including the NH sequences
# I also dropped HepA/USA/MA/0046/2018 
qc_filtered_metadata <- metadata %>%
  filter(Sample_Passed_QC == "Yes",
          Subgenotype == "IIIA",
         MASPHL_Sequencing_ID != "HepA/USA/MA/0046/2018")
colnames(qc_filtered_metadata)
nrow(qc_filtered_metadata)

```

```{r}
# plot transmissions rates by metadata variables
height <- 2.5
width <- 7

all_variables <- data.table(
  variable = c("vaccine", "gender", "homelessness", "unstablehousing", "recent_illicit", "illicit_ever", "recent_injection", "injection_ever", "nausea", "vomiting", "diarrhea", "jaundice", "IS_CASE_A_FOODHANDLER", "travel"),
  variable_name  = c("Vaccine", "Gender", "Homelessness", "Unstable Housing", "Illicit Drug Use (Recent)", "Illicit Drug Use (Ever)", "Injection Drug Use (Recent)", "Injection Drug Use (Ever)", "Nausea",   "Vomiting", "Diarrhea", "Jaundice", "Is Food Handler", "Travel History"),
  legend_name  = c("Vaccine", "Gender", "Experiencing Homelessness", "Unstable Housing", "Illicit Drug Use (Recent)", "Drug Use", "Injection Drug Use (Recent)", "Injection Drug Use (Ever)", "Nausea",   "Vomiting", "Diarrhea", "Jaundice", "Is Food Handler", "Travel History"),
  comparison_alt = c("No/Unknown_vs_Yes",  "Female_vs_Male", "No_vs_Yes", "Yes_vs_No", "No_vs_Yes", "No_vs_Yes", "No_vs_Yes", "No_vs_Yes", "Yes_vs_No","Yes_vs_No","Yes_vs_No","Yes_vs_No","No_vs_Yes", "No_vs_Yes"),
  comparison = c("Yes_vs_No/Unknown","Male_vs_Female", "Yes_vs_No", "No_vs_Yes", "Yes_vs_No", "Yes_vs_No", "Yes_vs_No", "Yes_vs_No", "No_vs_Yes","No_vs_Yes","No_vs_Yes","No_vs_Yes","Yes_vs_No", "Yes_vs_No"))

variables_fig3 <- all_variables %>% filter(variable %in% c('illicit_ever', 'homelessness'))
variables_to_assess <- variables_fig3

# blue #4B7BDC
# pink #F48FB1
# purple #6C5BD3
# yellow-orange #F2B134

for (col in variables_to_assess$variable) {
  cat("Column:", col, "\n")
  print(unique(qc_filtered_metadata[[col]]))
  print(sum(is.na(qc_filtered_metadata[[col]])))
}

juniper_results <- clade3a_juniper_pi_0.4_juniper
sampling_rate <- "0.4"

format_pvalue <- function(p) {
  if (p < 0.001) "p < 0.001" else paste("p = ", round(p, 3))
}

for (i in seq_len(nrow(variables_to_assess)))
{
  variable <- variables_to_assess$variable[i]
  variable_name <- variables_to_assess$variable_name[i]
  legend_name <- variables_to_assess$legend_name[i]
  this_comparison <- variables_to_assess$comparison[i]
  this_comparison_alt <- variables_to_assess$comparison_alt[i]
  
  print(variable)
  print(sampling_rate)

  if (variable=="homelessness") {
    colorway = c("#F2B134", "#6C5BD3") # orange and purple
    legend_pos = "left"
  } else if (variable == "illicit_ever") {
    colorway = c("#F48FB1", "#4B7BDC") # pink and blue
    legend_pos = "left"
  } else {
    colorway = c("#E3B861", "#4CB3A2") # default pale orange and teal
    legend_pos = "right"
  }
  
  transmission_distributions = transmission_by_variable_distributions(juniper_results, qc_filtered_metadata, variable=variable, name_var="MASPHL_Sequencing_ID", drop=c("missing", "Unknown"))

  p_values_table <- transmission_by_variable_comparison(juniper_results, qc_filtered_metadata, variable=variable, name_var="MASPHL_Sequencing_ID", method="b")

  if(variable == "IS_CASE_A_FOODHANDLER" || variable == "travel" || variable == "unstablehousing")
  {
    p_values_table <- transmission_by_variable_comparison(juniper_results, qc_filtered_metadata, variable=variable, name_var="MASPHL_Sequencing_ID", method="b", comparator="No")
  }
  
  probability_of_transmission_plot_p_value <- filter(p_values_table$p, comparison == this_comparison) %>% dplyr::pull(pval)

  if(length(probability_of_transmission_plot_p_value) == 0)
  {
    probability_of_transmission_plot_p_value <- filter(p_values_table$p, comparison == this_comparison_alt) %>% dplyr::pull(pval)
  }
  
  probability_of_transmission_plot_p_value_polished <- format_pvalue(probability_of_transmission_plot_p_value)
  print("probability of trasmission p value unpolished and polished")
  print(probability_of_transmission_plot_p_value)
  print(probability_of_transmission_plot_p_value_polished)
  probability_of_transmission_plot <- plot_transmission_by_variable(transmission_distributions, variable_name=variable_name, legend_name=legend_name, type="p", colors=colorway, title=probability_of_transmission_plot_p_value_polished)

  ggsave(paste(OUTPUT_DIR, "/fig3C_", variable, "_probability_of_transmission.jpg", sep=""), probability_of_transmission_plot, width=width, height=height)
  ggsave(paste(OUTPUT_DIR, "/fig3C_", variable, "_probability_of_transmission.pdf", sep=""), probability_of_transmission_plot, width=width, height=height)

  probability_of_transmission_means <- retrieve_means_for_transmission_by_variable(transmission_distributions, variable_name=variable_name, type="p")
  write.table(probability_of_transmission_means, paste(OUTPUT_DIR, "/fig3C_", variable, "_probability_of_transmission_means.txt", sep=""), sep = "\t", row.names = FALSE, quote = FALSE)

  number_transmissions_plot_p_value <- filter(p_values_table$n, comparison == this_comparison) %>% dplyr::pull(pval)

  if(length(number_transmissions_plot_p_value) == 0)
  {
    number_transmissions_plot_p_value <- filter(p_values_table$n, comparison == this_comparison_alt) %>% dplyr::pull(pval)
  }

  number_transmissions_plot_p_value_polished <- format_pvalue(number_transmissions_plot_p_value)
  print("num transmissions p value unpolshed and polished")
  print(number_transmissions_plot_p_value)
  print(number_transmissions_plot_p_value_polished)

  number_transmissions_plot <- plot_transmission_by_variable(transmission_distributions, variable_name=variable_name, legend_name=legend_name, type="n", colors=colorway, title=number_transmissions_plot_p_value_polished)
  ggsave(paste(OUTPUT_DIR, "/fig3C_", variable, "_number_transmissions.jpg", sep=""), number_transmissions_plot, width=width, height=height)
  ggsave(paste(OUTPUT_DIR, "/fig3C_", variable, "_number_transmissions.pdf", sep=""), number_transmissions_plot, width=width, height=height)

  number_transmissions_means <- retrieve_means_for_transmission_by_variable(transmission_distributions, variable_name=variable_name, type="n")
  write.table(number_transmissions_means, paste(OUTPUT_DIR, "/fig3C_", variable, "_number_transmissions_means.txt", sep=""), sep = "\t", row.names = FALSE, quote = FALSE)

  plot(number_transmissions_plot)
  plot(probability_of_transmission_plot)
}
```