---
title: "Figure 4"
execute:
  echo: false
---

**Figure 4. Comparison of HAV whole genome sequencing and subgenotyping approaches.**

**A)** Phylogenetic relationships among clade IIIA genomes from this study that were successfully sequenced and had known epidemiological links reconstructed using whole genome sequences (left) and the VP1-2B region only (right). Linkage group designation is indicated by matching letters (A-I) and like colors. 

**B)** Average sequencing coverage for samples (n=7) from the 2023-2024 outbreak period that produced full genome assemblies using each sequencing method. The VP1-2B region is marked by a dark gray bar. 

**C)** Percentage of reads per sample identified as HAV for each sequencing method. 

```{r}
rm(list = ls())
pkglist = c("ape", "ggtree", "ggplot2", "viridis", "dplyr", "stringr", "cowplot", "tidyr", "data.table", "gridExtra", "cowplot", "readxl", "tidyverse", "igraph")
lapply(pkglist, require, character.only = TRUE)
```

```{r}
OUTPUT_DIR=Sys.getenv("OUTPUT_DIR")
TR_TREE=Sys.getenv("TR_TREE")
WGS_TREE=Sys.getenv("WGS_TREE")
COVERAGE_XLSX=Sys.getenv("COVERAGE_XLSX") 
HAV_PANEL_SHEET=Sys.getenv("HAV_PANEL_SHEET")
TWIST_SHEET=Sys.getenv("TWIST_SHEET")
METAGENOMIC_COVERAGE=Sys.getenv("METAGENOMIC_COVERAGE")
TABLE_S1=Sys.getenv("TABLE_S1")
EPI_LINKAGE_SCRIPT=Sys.getenv("EPI_LINKAGE_SCRIPT")
TABLES1_SCRIPT=Sys.getenv("TABLES1_SCRIPT")

if (!file.exists(TABLE_S1) || file.size(TABLE_S1) == 0) {
  source(TABLES1_SCRIPT)
}

source(EPI_LINKAGE_SCRIPT)
```

```{r}
# 4A
outgroup <- "AB279732.1" # oldest genbank sequence from each clade is used as outgroup

# epi links here = all-vs-all QC-passing epi pairs derived from linkage groups
epi_links <- subset(combined, type == "epi" | type == "epi and juniper")
nrow(epi_links) #31

# function to root tree on outgroup and save
process_tree_with_outgroup <- function(tree_file,label,outgroup,OUTPUT_DIR,linkages) {
    tree <- read.tree(tree_file)

    # just in case - shouldn't be present but double check
    tree <- drop.tip(tree, "HepA/USA/MA/0046/2018")
    
    tree <- root(tree, outgroup = outgroup, resolve.root = TRUE)
    
    # is outgroup the furthest tip?
    distances <- node.depth.edgelength(tree)
    farthest_tip <- tree$tip.label[which.max(distances[1:Ntip(tree)])]
    
    if (farthest_tip != outgroup) {
        writeLines("Warning: farthest tip is not outgroup",con = file.path(OUTPUT_DIR,paste0(label, "_warning.txt")))
    }
    
    # drop outgroup
    tree <- drop.tip(tree, outgroup)
    # #write.tree(tree,file = file.path(OUTPUT_DIR,paste0(label,"_rerooted_with_outgroup_outgroup_removed.txt")))
    
    # annotate tree with linkages
    tree_data <- data.frame(label = tree$tip.label)
    tree_data <- merge(tree_data, linkages,by.x = "label",by.y = "strain",all.x = TRUE)

    if (label == "TR") {
        text_offset <- 0.00000003
        scale_width <- 0.0000001
        plot_title <- "VP1-2B"
    } else if (label == "WGS") {
        text_offset <- 0.00001
        scale_width <- 0.0001
        plot_title <- "WGS"
    } else {
        text_offset <- 0.00001
        scale_width <- 0.0001
        plot_title <- label
    }
    
    # plot tree
    tree_with_linkages <- ggtree(tree) %<+% tree_data +
        geom_tippoint(aes(color = linkage_group), size = 3) +
        geom_tiplab(aes(label = linkage_group),
                    size = 3,
                    color = "black",
                    offset = text_offset) +
        scale_color_brewer(palette = "Spectral", na.translate = FALSE) +
        theme_tree2() +
        theme(legend.position = "none",
            axis.line.x = element_blank(),
            axis.ticks.x = element_blank(),
            axis.text.x = element_blank(),
            axis.title.x = element_blank()) +
        geom_treescale(x = 0,
                    y = 0,
                    width = scale_width,
                    fontsize = 2) +
        ggtitle(plot_title)
    
    # add root branch in
    root_y <- tree_with_linkages$data$y[
        tree_with_linkages$data$node ==
        getMRCA(tree, tree$tip.label)
    ]
    
    tree_with_linkages <- tree_with_linkages +
        annotate("segment",
                x = -2 * text_offset,
                xend = 0,
                y = root_y,
                yend = root_y)
    
    plot(tree_with_linkages)
    ggsave(paste(OUTPUT_DIR, "/Fig4A_", label, ".jpg", sep=""), tree_with_linkages, width=4, height=4)
    ggsave(paste(OUTPUT_DIR, "/Fig4A_", label,".pdf", sep=""), tree_with_linkages, width=4, height=4)
}

process_tree_with_outgroup(TR_TREE, "TR", outgroup, OUTPUT_DIR, linkage_groups)

process_tree_with_outgroup(WGS_TREE, "WGS", outgroup, OUTPUT_DIR, linkage_groups)

```

```{r}
# 4B

HAV_Panel_Coverage <- read_excel(COVERAGE_XLSX, sheet = HAV_PANEL_SHEET)
Twist_Coverage <- read_excel(COVERAGE_XLSX, sheet = TWIST_SHEET)
Metagenomic_Coverage <- read_excel(COVERAGE_XLSX, sheet = METAGENOMIC_COVERAGE)


## HAV Panel
p1 = ggplot(HAV_Panel_Coverage, aes(x=`Position on genome`, y=`Average  Read Depth`)) +  
  geom_area(alpha=0.4 , size=.5, colour="black", fill="#999999") +
  scale_fill_viridis(discrete = T) + 
  theme_classic() +
  theme(axis.title = element_text(size = 12, face = "bold")) +
  ggtitle("HAV Specific Panel") +
  theme(plot.title = element_text(size = 12, face = "bold")) +
  labs(x = "Genome Position", y = "") +
  scale_y_continuous(expand = c(0,0),limits = c(0,1100)) +
  scale_x_continuous(expand = c(0,0), limits = c(0,7500), breaks = c(0,500, 1000, 1500, 2000, 2500, 3000, 3500, 4000, 4500, 5000, 5500, 6000, 6500, 7000, 7500)) +
  theme(axis.text.x = element_text(angle=90, vjust=.5, hjust=1)) +
  geom_segment(aes(x = 2897, y = 5, xend = 3288), size=7, color="#666666")


## Twist Panel
p2 = ggplot(Twist_Coverage, aes(x=`Position on genome`, y=`Average  Read Depth`)) +  
  geom_area(alpha=0.4 , size=.5, colour="black", fill="#999999") +
  scale_fill_viridis(discrete = T) + 
  theme_classic() +
  theme(axis.title = element_text(size = 12, face = "bold")) +
  theme(plot.title = element_text(size = 12, face = "bold")) +
  ggtitle("Twist Panel") +
  labs(x = "", y = "Average Read Depth") +
  scale_y_continuous(expand = c(0,0),limits = c(0,1100)) +
  scale_x_continuous(expand = c(0,0), limits = c(0,7500), breaks = c(0,500, 1000, 1500, 2000, 2500, 3000, 3500, 4000, 4500, 5000, 5500, 6000, 6500, 7000, 7500)) +
  theme(axis.text.x = element_text(angle=90, vjust=.5, hjust=1)) +
  geom_segment(aes(x = 2897, y = 5, xend = 3288), size=7, color="#666666")

## Metagenomic
p3 = ggplot(Metagenomic_Coverage, aes(x=`Position on genome`, y=`Average  Read Depth`)) +  
  geom_area(alpha=0.4 , size=.5, colour="black", fill="#999999") +
  scale_fill_viridis(discrete = T) + 
  theme_classic() +
  theme(axis.title = element_text(size = 12, face = "bold")) +
  theme(plot.title = element_text(size = 12, face = "bold")) +
  ggtitle("Metagenomic") +
  labs(x = "", y = "") +
  scale_y_continuous(expand = c(0,0),limits = c(0,1100)) +
  scale_x_continuous(expand = c(0,0), limits = c(0,7500), breaks = c(0,500, 1000, 1500, 2000, 2500, 3000, 3500, 4000, 4500, 5000, 5500, 6000, 6500, 7000, 7500)) +
  theme(axis.text.x = element_text(angle=90, vjust=.5, hjust=1)) +
  geom_segment(aes(x = 2897, y = 5, xend = 3288), size=7, color="#666666")


##### Combined plot
fig4B <- grid.arrange(p3, p2, p1, ncol=1)
plot(fig4B)
ggsave(paste(OUTPUT_DIR, "/Fig4B", ".jpg", sep=""), fig4B)
ggsave(paste(OUTPUT_DIR, "/Fig4B", ".pdf", sep=""), fig4B)
```

```{r}
# 4C
## read table S1
Table_S1 <- read_csv(TABLE_S1)

#HAV Panel Plot
df_HAV_Panel = select(filter(Table_S1,`Sample ID`=="HepA/USA/MA/0354/2023"|`Sample ID`=="HepA/USA/MA/0355/2023"|`Sample ID`=="HepA/USA/MA/0355/2023"|`Sample ID`=="HepA/USA/MA/0356/2023"|`Sample ID`=="HepA/USA/MA/0357/2023"|`Sample ID`=="HepA/USA/MA/0359/2023"|`Sample ID`=="HepA/USA/MA/0360/2023"|`Sample ID`=="HepA/USA/MA/0361/2023"|`Sample ID`=="HepA/USA/MA/0362/2023"|`Sample ID`=="HepA/USA/MA/0363/2024"|`Sample ID`=="HepA/USA/MA/0364/2024"|`Sample ID`=="HepA/USA/MA/0367/2024"|`Sample ID`=="HepA/USA/MA/0369/2024"|`Sample ID`=="HepA/USA/MA/0373/2024"|`Sample ID`=="HepA/USA/MA/0376/2024"), c('Sample ID','Percent of Unclassified Reads HAV Panel','Percent of Other Viral Reads HAV Panel','Percent of HAV Reads HAV Panel'))
df_long_HAV_Panel = melt(as.data.table(df_HAV_Panel))

p3 = ggplot(df_long_HAV_Panel, aes(x = `Sample ID`, y = value, fill = variable)) + 
  geom_bar(stat="identity") +
  theme_classic() +
  theme(axis.title = element_text(size = 12, face = "bold")) +
  theme(axis.text.x = element_text(angle=90, vjust=.5, hjust=1)) +
  theme(plot.title = element_text(size = 12, face = "bold")) +
  scale_fill_manual(name="Read Classification",labels=c("Unclassified","Other Viral", "Hepatovirus A"), values=c("#E5E5E5","#999999","#666666")) +
  #labs(x = "Sample", y = "Percent of total reads") +
  labs(y ="", x = "") +
  scale_y_continuous(expand = c(0,0))+ 
  ggtitle("HAV Specific Panel") +
  theme(legend.position="none")

p3
#Twist Panel
df_Twist = select(filter(Table_S1,`Sample ID`=="HepA/USA/MA/0354/2023"|`Sample ID`=="HepA/USA/MA/0355/2023"|`Sample ID`=="HepA/USA/MA/0355/2023"|`Sample ID`=="HepA/USA/MA/0356/2023"|`Sample ID`=="HepA/USA/MA/0357/2023"|`Sample ID`=="HepA/USA/MA/0359/2023"|`Sample ID`=="HepA/USA/MA/0360/2023"|`Sample ID`=="HepA/USA/MA/0361/2023"|`Sample ID`=="HepA/USA/MA/0362/2023"|`Sample ID`=="HepA/USA/MA/0363/2024"|`Sample ID`=="HepA/USA/MA/0364/2024"|`Sample ID`=="HepA/USA/MA/0367/2024"|`Sample ID`=="HepA/USA/MA/0369/2024"|`Sample ID`=="HepA/USA/MA/0373/2024"|`Sample ID`=="HepA/USA/MA/0376/2024"), c('Sample ID','Percent of Unclassified Reads Twist','Percent of Other Viral Reads Twist','Percent of HAV Reads Twist'))
df_Twist_long = melt(as.data.table(df_Twist))

p2 = ggplot(df_Twist_long, aes(x = `Sample ID`, y = value, fill = variable)) + 
  geom_bar(stat="identity") +
  theme_classic() +
  theme(plot.title = element_text(size = 12, face = "bold")) +
  theme(axis.title = element_text(size = 12, face = "bold")) +
  theme(axis.text.x = element_text(angle=90, vjust=.5, hjust=1)) +
  scale_fill_manual(name="Read Classification",labels=c("Unclassified", "Other Viral", "Hepatovirus A"), values=c("#E5E5E5","#999999","#666666")) +
  #labs(x = "Sample", y = "Percent of total reads") +
  labs(y ="", x = "") +
  scale_y_continuous(expand = c(0,0)) + 
  ggtitle("Twist Panel")+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

#Metagenomic plot
df_Meta = select(filter(Table_S1,`Sample ID`=="HepA/USA/MA/0354/2023"|`Sample ID`=="HepA/USA/MA/0355/2023"|`Sample ID`=="HepA/USA/MA/0355/2023"|`Sample ID`=="HepA/USA/MA/0356/2023"|`Sample ID`=="HepA/USA/MA/0357/2023"|`Sample ID`=="HepA/USA/MA/0359/2023"|`Sample ID`=="HepA/USA/MA/0360/2023"|`Sample ID`=="HepA/USA/MA/0361/2023"|`Sample ID`=="HepA/USA/MA/0362/2023"|`Sample ID`=="HepA/USA/MA/0363/2024"|`Sample ID`=="HepA/USA/MA/0364/2024"|`Sample ID`=="HepA/USA/MA/0367/2024"|`Sample ID`=="HepA/USA/MA/0369/2024"|`Sample ID`=="HepA/USA/MA/0373/2024"|`Sample ID`=="HepA/USA/MA/0376/2024"), c('Sample ID','Percent of Unclassified Reads Metagenomic','Percent of Other Viral Reads Metagenomic','Percent of HAV Reads Metagenomic'))
df_Meta_long = melt(as.data.table(df_Meta))

p1 = ggplot(df_Meta_long, aes(x = `Sample ID`, y = value, fill = variable)) + 
  geom_bar(stat="identity") +
  theme_classic() +
  scale_fill_manual(name="Read Classification",labels=c("Unclassified", "Other Viral", "Hepatovirus A"), values=c("#E5E5E5","#999999","#666666")) +
  theme(axis.text.x = element_text(angle=90, vjust=.5, hjust=1)) +
  theme(axis.title = element_text(size = 12, face = "bold")) +
  theme(plot.title = element_text(size = 12, face = "bold")) +
  #labs(x = "Sample", y = "Percent of total reads") +
  labs(y ="", x = "") +
  scale_y_continuous(expand = c(0,0)) +
  ggtitle("Metagenomic") +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  theme(legend.position="none")

#Create combined plot
combined_plot = plot_grid(p1, p2, p3, align = "v", nrow = 3, rel_heights = c(1/4, 1/4, 8/17))

final_plot2 <- ggdraw(combined_plot) +
  draw_label("Sample", x = 0.5, y = 0, vjust = -0.5, hjust= 1.2, angle = 0, size = 13.5, fontface="bold") +
  draw_label("Percent of Reads", x = 0, y = 0.5, vjust = 1.5, angle = 90, size = 13.5, fontface="bold")

plot(final_plot2)

ggsave(paste(OUTPUT_DIR, "/Fig4C", ".jpg", sep=""), final_plot2)
ggsave(paste(OUTPUT_DIR, "/Fig4C", ".pdf", sep=""), final_plot2)
```