---
title: "Figure S5"
execute:
  echo: false
---

**Figure S5. Association between geographic distance and SNP distance for MA IIIA sequences.** Association between geographic distance and genomic distance, as measured by SNP distance, for MA IIIA cases from 2018-2020. 

**A)** Geographic distance as measured between city of residence recorded for each patient sample. 

**B)** Geographic distance as measured between city where a provider ordered a diagnostic test for each patient sample.  

```{r}
rm(list = ls())
pkglist = c("gridExtra", "dplyr", "ggplot2","ggpmisc", "ape", "cowplot","tibble", "data.table")
lapply(pkglist, require, character.only = TRUE)
```

```{r}
OUTPUT_DIR=Sys.getenv("OUTPUT_DIR")
METADATA_FILE=Sys.getenv("METADATA_FILE")
CITIES_LAT_LONG_FILE=Sys.getenv("CITIES_LAT_LONG_FILE")
RENAMED_ALL_PAIRS_SNP_DIST=Sys.getenv("RENAMED_ALL_PAIRS_SNP_DIST")

```

```{r}
# read in metadata and associated files
metadata <- read.table(METADATA_FILE, sep=",", header=TRUE, stringsAsFactors=FALSE)

# drop HepA/USA/MA/0046/2018
metadata <- metadata[metadata$MASPHL_Sequencing_ID != "HepA/USA/MA/0046/2018", ]

metadata$official_city <- toupper(metadata$official_city) #unnecessary?
metadata$diagnostic_provider_city <- toupper(metadata$diagnostic_provider_city) # def necessary for this column
cities_lat_long <- read.table(CITIES_LAT_LONG_FILE, sep="\t", header=TRUE, stringsAsFactors=TRUE)
```

```{r}
# subset all-vs-all SNP distances to the three subsets
all_pairs <- read.table(RENAMED_ALL_PAIRS_SNP_DIST, sep="\t", header=TRUE, stringsAsFactors=TRUE)
nrow(all_pairs)

# drop anything with HepA/USA/MA/0046/2018 from all_pairs
all_pairs <- all_pairs[all_pairs$sequence1 != "HepA/USA/MA/0046/2018" & 
                       all_pairs$sequence2 != "HepA/USA/MA/0046/2018", ]

# Get IDs that satisfy all conditions simultaneously (from MA, passed QC, subgenotype IIIA)
valid_ids <- metadata$MASPHL_Sequencing_ID[
  metadata$State.of.Residence == "MA" &
  metadata$Sample_Passed_QC == "Yes" &
  metadata$Subgenotype == "IIIA" &
  grepl("(2018|2019|2020)$", metadata$MASPHL_Sequencing_ID)]


# Filter pairs where BOTH samples meet all criteria
all_pairs <- all_pairs[all_pairs$sequence1 %in% valid_ids & 
                       all_pairs$sequence2 %in% valid_ids, ]
nrow(all_pairs) 

# add homelessness, official_city for both sequence1 and sequence2
all_pairs <- all_pairs %>%
  left_join(metadata %>% select(sequence1 = MASPHL_Sequencing_ID, sequence1_homelessness = homelessness, sequence1_official_city = official_city, sequence1_diagnostic_provider_city = diagnostic_provider_city),
            by = "sequence1") %>%
  left_join(metadata %>% select(sequence2 = MASPHL_Sequencing_ID, sequence2_homelessness = homelessness, sequence2_official_city = official_city, sequence2_diagnostic_provider_city = diagnostic_provider_city),
            by = "sequence2")

# add latitude and longitude for offical city of residence
all_pairs <- all_pairs %>%
  left_join(cities_lat_long %>% select(sequence1_official_city = city, sequence1_longitude = longitude, sequence1_latitude=latitude),
            by = "sequence1_official_city") %>%
  left_join(cities_lat_long %>% select(sequence2_official_city = city, sequence2_longitude = longitude, sequence2_latitude=latitude),
            by = "sequence2_official_city")

# calculate geographic distances between MA cities
all_pairs$lat_diff <- abs(all_pairs$sequence1_latitude - all_pairs$sequence2_latitude)
all_pairs$long_diff <- abs(all_pairs$sequence1_longitude - all_pairs$sequence2_longitude)
all_pairs$geographic_distance <- sqrt((111*all_pairs$lat_diff)^2 + (83*all_pairs$long_diff)^2)

# now add ordering provider cities
all_pairs <- all_pairs %>%
  left_join(cities_lat_long %>% select(sequence1_diagnostic_provider_city = city, sequence1_ordering_provider_longitude = longitude, sequence1_ordering_provider_latitude=latitude),
            by = "sequence1_diagnostic_provider_city") %>%
  left_join(cities_lat_long %>% select(sequence2_diagnostic_provider_city = city, sequence2_ordering_provider_longitude = longitude, sequence2_ordering_provider_latitude=latitude),
            by = "sequence2_diagnostic_provider_city")

all_pairs$ordering_provider_lat_diff <- abs(all_pairs$sequence1_ordering_provider_latitude - all_pairs$sequence2_ordering_provider_latitude)
all_pairs$ordering_provider_long_diff <- abs(all_pairs$sequence1_ordering_provider_longitude - all_pairs$sequence2_ordering_provider_longitude)
all_pairs$ordering_provider_geographic_distance <- sqrt((111*all_pairs$ordering_provider_lat_diff)^2 + (83*all_pairs$ordering_provider_long_diff)^2)

all_pairs_subset <- all_pairs
nrow(all_pairs_subset)
```

```{r}
# 5A 
# generate geographic vs snp distance plots for official city of residence
model <- lm(geographic_distance ~ snp_distance, data = all_pairs_subset)
r_squared <- summary(model)$r.squared
r_squared_polished <- paste("R² = ", round(r_squared, 3), sep="")
geographic_vs_snp_distance_IIIA_2018_2020_plot <- ggplot(data = all_pairs_subset, aes(x = snp_distance, y = geographic_distance)) +
    geom_point(alpha = 0.3) +
    geom_smooth(method = "lm", se = FALSE, color = "black") +
    xlab("SNP Distance") + ylab("Geographic Distance (km)") +
    ggtitle(r_squared_polished) +
    theme_minimal() +
    theme(panel.grid=element_blank(), axis.line=element_line(), axis.ticks=element_line())

# calculate geographic distance for SNP distance 0 vs. >0
identical_genomes_cutoff = 0
identical_genomes_label <- "Identical Genomes"
nonidentical_genomes_label <- "SNP Distance >0"
all_pairs_subset$identical_genomes <- NA
all_pairs_subset$identical_genomes[all_pairs_subset$snp_distance <= identical_genomes_cutoff] <- identical_genomes_label
all_pairs_subset$identical_genomes[all_pairs_subset$snp_distance > identical_genomes_cutoff] <- nonidentical_genomes_label
all_pairs_subset$identical_genomes <- factor(all_pairs_subset$identical_genomes, levels = c(identical_genomes_label, nonidentical_genomes_label))

## Wilcoxon rank-sum test
p_value <- wilcox.test(geographic_distance ~ identical_genomes, data = all_pairs_subset)$p.value
p_value_polished <- if (p_value < 0.001) "p < 0.001" else paste("p =", signif(p_value, 3))

# plot geographic distance for SNP distance 0 vs. >0
geographic_distance_identical_genomes_plot <- ggplot(data = subset(all_pairs_subset, !is.na(identical_genomes)), aes(x = identical_genomes, y = geographic_distance)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.2, alpha = 0.3, color = "black") +
    theme_bw() +
    theme(panel.grid=element_blank(), axis.ticks.x=element_blank()) +
    ylab("Geographic Distance (km)") +
    xlab("") +
    ggtitle(p_value_polished)

# arrange and save final figure with four panels
final_plot <- plot_grid(geographic_vs_snp_distance_IIIA_2018_2020_plot, geographic_distance_identical_genomes_plot,
rel_widths = c(1, 0.6), align = 'v', axis = 'l', labels = NULL)
plot(final_plot)  

print(summary(model))

ggsave(paste(OUTPUT_DIR, "figS5A_official_city_of_residence.png", sep="/"), final_plot, width=9, height=4.5)
ggsave(paste(OUTPUT_DIR, "figS5A_official_city_of_residence.pdf", sep="/"), final_plot, width=9, height=4.5)
```

```{r}
# 5B - generate geographic vs snp distance plots for ordering provider city
model <- lm(ordering_provider_geographic_distance ~ snp_distance, data = all_pairs_subset)
r_squared <- summary(model)$r.squared
r_squared_polished <- paste("R² = ", round(r_squared, 3), sep="")
print(r_squared_polished)

geographic_vs_snp_distance_IIIA_2018_2020_plot <- ggplot(data = all_pairs_subset, aes(x = snp_distance, y = ordering_provider_geographic_distance)) +
    geom_point(alpha = 0.3) +
    geom_smooth(method = "lm", se = FALSE, color = "black") +
    xlab("SNP Distance") + ylab("Geographic Distance (km)") +
    ggtitle(r_squared_polished) +
    theme_minimal() +
    theme(panel.grid=element_blank(), axis.line=element_line(), axis.ticks=element_line())

# calculate geographic distance for SNP distance 0 vs. >0
identical_genomes_cutoff = 0
identical_genomes_label <- "Identical Genomes"
nonidentical_genomes_label <- "SNP Distance >0"
all_pairs_subset$identical_genomes <- NA
all_pairs_subset$identical_genomes[all_pairs_subset$snp_distance <= identical_genomes_cutoff] <- identical_genomes_label
all_pairs_subset$identical_genomes[all_pairs_subset$snp_distance > identical_genomes_cutoff] <- nonidentical_genomes_label
all_pairs_subset$identical_genomes <- factor(all_pairs_subset$identical_genomes, levels = c(identical_genomes_label, nonidentical_genomes_label))

# Wilcoxon rank-sum test
p_value <- wilcox.test(ordering_provider_geographic_distance ~ identical_genomes, data = all_pairs_subset)$p.value
p_value_polished <- if (p_value < 0.001) "p < 0.001" else paste("p =", signif(p_value, 3))
#p_value_polished <- paste("p =", formatC(p_value, format = "g", digits = 3))

# plot geographic distance for SNP distance 0 vs. >0
geographic_distance_identical_genomes_plot <- ggplot(data = subset(all_pairs_subset, !is.na(identical_genomes)), aes(x = identical_genomes, y = ordering_provider_geographic_distance)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.2, alpha = 0.3, color = "black") +
    theme_bw() +
    theme(panel.grid=element_blank(), axis.ticks.x=element_blank()) +
    ylab("Geographic Distance (km)") +
    xlab("") +
    ggtitle(p_value_polished)

# arrange and save final figure with four panels
final_plot <- plot_grid(geographic_vs_snp_distance_IIIA_2018_2020_plot, geographic_distance_identical_genomes_plot,
rel_widths = c(1, 0.6), align = 'v', axis = 'l', labels = NULL)
plot(final_plot)

print(summary(model))

ggsave(paste(OUTPUT_DIR, "figS5B_ordering_provider_city.png", sep="/"), final_plot, width=9, height=4.5)
ggsave(paste(OUTPUT_DIR, "figS5B_ordering_provider_city.pdf", sep="/"), final_plot, width=9, height=4.5)

```